Simulating High-Dimensional Multivariate Data

### 2020-11-06 010-background.Rmd
2nd. Draft. Revise background

The article presents several algorithms to work with high-dimesional multivariate data, but all `bigsimr` algorithms were originally designed to support a single task:
to generate random vectors drawn from multivariate probability distribution with given marginal distributions and component-wise dependency metrics.
Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and hetereogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, and $d$ can be very large.

When designing this methodology, we developed the following properties to guide our effort.
We divide the properties into two categories: (1) basic properties (**BP**) and "scaleability" properties (**SP**). 
The BPs are adapted from an existing criteria due to @Nik13a.
Our simulation strategy should allow:

@@

To fix ideas and provide example applications enabled via `bigsimr`, the next section describes a motivating data set that originally inspired the authors` interest in developing this methodology.

@@

Simulating high-dimensional, non-normal, correleated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
This results in count data corresponding to over 20,000 genes (coding genomic regions) or even over 77,000 isoforms when alternating spliced mRNA are counted (for human derived samples).
Yet due to inherent biological processes, gene expression data exhibits correlation (co-expression) across genes [@BE07; @Schissler2018].

```{r ch010-myProb, include=FALSE, cache = TRUE}
myProb <- 0.99
```

@@

Specifically, we'll illustrate our methodology using the Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgements).
For simplicity, we begin by filtering to retain the top `r (1-myProb)*100`% highest expressing (in terms of median expression) genes of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
This results in $`r choose(d,2)`$ pairwise dependencies among the bivariate marginals.
To illustrate our methodology's flexible/robust simulation scheme and better align with the actual data-generating process, we remove the statistical adjustment for read alignment ambiguity (RNA-seq by Expectation Maximization; RSEM; @Li2011c) by simple rounding to integer values.
Table \@ref(tab:ch010-realDataTab) displays counts for three selected genes for the first five patients' breast tumor samples.

@@

Figure \@ref(fig:ch010-realDataFig) plots the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).

```{r ch010-realDataFig, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's $rho$ for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
## Describe and plot real data
## NEED TO CHANGE TO SPEARMANS
GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

@@

As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. 
For non-normal marginals with monotone correlation patterns, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships (@MK01).
Alternatively, in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparameteric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
Formally, define

\begin{equation}
(\#eq:spearman)
\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
\end{equation}

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. 
Spearman's $\rho_S$ has an appealing correspondence as the Pearson's correlaiton coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.

@@

Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
As we will describe more fully in the [Algorithms](#algorithms) section, this property is preferred in our scheme to satistify properties SP1 and SP2, above.

@@

Also we'll note that for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$.
Let

@@

*Marginal-dependent bivariate correlation bounds*. 
Given two marginal distributions, $\rho_P$ is not free to vary the entire range of possible correlations $[-1,1]$.
The so-called *Frechet-Hoeffding bounds* are well-studied (for example, see @Nelsen2007 and @BF17).
This situation gives strict restraints on the possible bounds and cannot be overcome through algorithm design.
In general the bounds are given by

@@

There is a strong connection of our simulation strategy to Gaussian **copulas** (for a technical introduction see @Nelsen2007).
A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
This definition provides a powerful, natural way characterize joint probability structure.
Consequently, the study of copulas is an important and active area of statistical theory and practice.

>>>>

The `bigsimr` `R` package provides multiple algorithms to work with high-dimesional multivariate data, but all these algorithms were originally designed to support a single task:
to generate random vectors drawn from multivariate probability distribution with given marginal distributions and dependency metrics.
Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and hetereogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful times.

When designing this methodology, we developed the following properties to guide our effort.
We divide the properties into two categories: (1) basic properties (BP) and "scaleability" properties (SP). 
The BPs are adapted from an existing criteria due to @Nik13a.
Our simulation strategy should allow:

@@

To fix ideas and provide examples applications enabled via `bigsimr`, the next section describes a motivating data set that originally inspired the authors' interest in developing this methodology.

@@

Simulating high-dimensional, non-normal, correleated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].

```{r ch010-myProb, include=FALSE, cache = TRUE}
myProb <- 0.95
```

@@

We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgements).
For simplicity, we only consider high expressing genes.
In turn, we begin by filtering to retain the top `r (1-myProb)*100`% highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c's RNA-seq by Expectation Maximization (RSEM) values to counts.
Table \@ref(tab:ch010-realDataTab) displays counts for three selected high-expressing genes for the first five patients' breast tumor samples.

@@

To help visualize the bivariate relationships for these three selected genes across all patients, Figure \@ref(fig:ch010-realDataFig) plots the marginal distributions and estimated Spearman's correlations (see Equation \@ref(eq:spearman) below).

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's $\rho_s$ for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
## Describe and plot real data
## NEED TO CHANGE TO SPEARMANS
GGally::ggpairs(data = as.data.frame(brca[ ,exampleGenes] ))
```

@@

As @MB13 and @MK01 discuss, for a bivariate normal $(X,Y)$ random vector, the Pearson correlation completely describes the dependency between the components. 
For non-normal marginals with monotone correlation patterns, $\rho_P$ suffers some drawbacks and may mislead or fail to capture important relationships (@MK01).
Alternatively in these settings, analysts often prefer rank-based correlation measures to describe the degree of monotonic association.

Two nonparameteric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
Define

\begin{equation}
(\#eq:spearman)
\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right],
\end{equation}

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. 
Spearman's $\rho_S$ has an appealing correspondence as the Pearson's correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.

@@

Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).

@@

Also, for a discrete random variable $Y_i$, some care must be taken to define the quantile function $F_{i}^{-1}$.
Let

@@

*Marginal-dependent bivariate correlation bounds*. 
Given two marginal distributions, $\rho_P$ is not free to vary the entire range of possible correlations $[-1,1]$.
The so-called *Frechet-Hoeffding bounds* are well-studied (for example, see @Nelsen2007 and @BF17).
This situation gives strict restraints on the possible correlations and cannot be overcome through algorithm design.
In general, the bounds are given by

@@

There is a strong connection of our simulation strategy to Gaussian **copulas** (see @Nelsen2007 for a technical introduction).
A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
This definition provides a powerful, natural way characterize joint probability structure.
Consequently, the study of copulas is an important and active area of statistical theory and practice.

### 2020-11-06 index.Rmd
2nd. Draft. Revise background

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author: "Alexander D. Knudson, Tomasz J. Kozubowski, Anna K. Panorska, Juli Petereit, and Alfred G. Schissler"
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d>1000$.
---

@@

```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(tidyverse)
library(knitr)
knitr::opts_chunk$set(eval = TRUE, echo = FALSE, cache = TRUE)
library(rmarkdown)
library(bookdown)
## setwd('~/article_bigsimr')
set.seed(09212020)
```

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author: "Alexander D. Knudson, Tomasz J. Kozubowski, Anna K. Panorska, Juli Petereit, and Alfred G. Schissler"
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$.
---

@@

```{r indexSetup, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide", cache=F}
library(tidyverse)
library(knitr)
knitr::opts_chunk$set(eval = TRUE, echo = FALSE, cache = TRUE)
library(rmarkdown)
library(bookdown)
## setwd('~/article_bigsimr')
set.seed(09212020)
```
